The demographic costs of sexually antagonistic selection in partially selfing populations

### 2022-01-03 doc/SA-Hermaphrodites-wDemography-R1.tex
more minor tweaks

%To persist in the long term, a population must be able to adapt to its environment. Yet, the process of adaptation can be impeded by a variety of genetic and environmental factors, such as deleterious mutations \citep{Haldane1937}, \textcolor{blue}{rapid, unpredictable, or extreme changes} in environmental conditions \citep{Maynard-Smith1976, LandeShannon1996, OrrUnckless2008,Lyberger-etal-2021}, maladaptive gene-flow \citep{KirkpatrickBarton1997, BolnickNosil2007}, and genetic constraints \citep{ConnallonHall2018, MatthewsConnallon2019}. Genetic constraints on adaptation can arise as a consequence of conflicting selection and gene-flow between different classes of individuals, or more generally, when \textcolor{blue}{different fitness components are negatively correlated} \citep{CharlesworthHughes2000, ConnallonHall2018}. 
\textcolor{blue}{A population's ability to adapt to its environment and persist in the long term depends upon the nature of the genetic variation that it harbors \citep{Fisher1930, Lewontin1975}. Yet, species exist in a complex world where the process of adaptation can be impeded by conflicting selection and gene-flow between different classes of individuals, or more generally, when different fitness components are negatively correlated \citep{Prout2000, CharlesworthHughes2000, ConnallonHall2018}.}

@@

Sexually antagonistic selection (abbreviated SA hereafter) \textcolor{blue}{is an important form of conflicting selection that occurs when beneficial alleles for one sex are deleterious when expressed in the other. Likewise, alleles with opposing fitness effects through male and female sex functions can cause SA selection in hermaphrodite populations, where individuals' fitness is determined jointly by their maternal and paternal reproductive success \citep{LloydWebb1986, WebbLloyd1986, Abbott2011, JordanConnallon2014}. SA selection appears to be} a common feature of sexually reproducing populations, and is thought to contribute to the maintenance of genetic variation and the evolution of sexual dimorphism and sex-related traits \citep{Kidwell1977, Lande1980, Rice1992, Charlesworth1999, RiceChippindale2001, ConnallonClark2012, BondurianskyChenoweth2009, Olito2019}. \textcolor{blue}{For instance, phenotypic selection studies commonly find sex-differences in directional selection \citep[e.g.,][]{DeLisle-etal-2018,Lewis-etal-2011,CoxCalsbeek2009,SinghPunzalan2018}, implying that many genetic variants underlying quantitative traits have SA fitness effects. Likewise, quantitative genetic studies frequently find negative cross-sex genetic correlations for fitness (\citealt{Chippendale-etal-2001,Delph-etal-2011}; reviewed in \citealt{ConnallonMatthews2019}), and polygenic signals of SA selection are evident in recent population genetic analyses of large genomic datasets \citep{RuzickaESEB2020, RuzickaConnallon2022}. Overall, both theory and current empirical data suggest that there is ample scope for the maintenance of SA \textcolor{blue}{genetic variation} in both dioecious \textcolor{blue}{(separate-sexed)} and hermaphrodite populations \citep{BondurianskyChenoweth2009, Abbott2011, RuzickaESEB2020, WangBarrett2020}.}

@@

\textcolor{blue}{While significant efforts have been made to predict when SA polymorphisms should be maintained and to detect SA genetic variation from empirical data, less attention has been paid to the demographic consequences of SA selection \citep{MatthewsConnallon2019}}. Most population genetic theory, including SA selection theory, models the relative fitness of genotypes in populations of constant (and often infinite) size. In reality, these genotypic fitnesses emerge from myriad processes acting throughout the life-history of individuals \textcolor{blue}{which link genotype distributions to population dynamics (e.g., \citealt {johnston2013life, merot2020balancing}). For over 50 years, the field of eco-evolutionary dynamics has sought to jointly model the genetic and demographic processes involved in evolutionary change. Theoretical models combining population genetics with density-dependent population growth \citep{roughgarden1971density} and stage-structured populations (consolidated in \citealt{Charlesworth_1994}) paved the way toward understanding the demographic consequences of allele frequency changes due to selection. Various extensions of these theories have been made to study life-history evolution \citep{orive1995senescence, orive2001somatic}, spatial structure \citep{ronce2000kin}, kin selection \citep{rousset2004genetic}, and the evolution of quantitative traits \citep[e.g.,][]{coulson2006putting, barfield2011evolution, childs2016evolution, orive2017effects}. This growing body of eco-evolutionary theory highlights that demographic processes must be taken into account explicitly in order to fully understand how genetic variation for fitness is shaped and maintained in natural populations.}

@@

\textcolor{blue}{One particularly important limitation of our analysis is that we modeled density-independent growth. Of course, in natural populations density-dependent processes like resource competition or competition for breeding sites will often prevent populations from growing (or shrinking) exponentially. In \ref{App:DensDep} we extend our model to include a simple form of density-dependence that acts on the survival of all life history stages equally, and reanalyse our case study of \textit{M. guttatus}. The density-dependent version of the model predicts equilibrium population densities rather than population intrinsic growth rates (see fig.~ \ref{fig:mimulusFig_dd}). This brief example shows that the shape of the polymorphic region is unchanged by this kind of density-dependence, and that the shape of the population density surface in fig.~\ref{fig:mimulusFig_dd} is very similar to that for population growth rate in fig.~\ref{fig:mimulusFig}. The similarity between the density-independent and -dependent model predictions underscores an important point: density-dependence that acts age-independently does not alter selection gradients \citep{mylius1995evolutionarily, caswell2017}.}

\textcolor{blue}{In many natural populations, however, density-dependence will not have the same effect on the survival of individuals of all ages or developmental stages. For example, competition for breeding sites will cause a reduction in fecundity rather than survival. Likewise, in plants density-dependent competition for soil nutrients will likely influence seed germination rates differently than seedling growth rates or flower/fruit production in adults \citep{AntonovicsLevin1980}. These types of age- or stage-specific density-dependence will change the selection gradients, and therefore the shape of polymorphic parameter space as well as the equilibrium population densities predicted by the model. \citet{de2020matrix} discuss some of the possible dynamical outcomes of a density-dependent genetic model for \textit{Tribolium} seed beetles but a thorough exploration of the consequences of more complicated forms of density-dependence for the maintenance of sexually antagonistic polymorphisms remains an important avenue for future work.}

@@

\textcolor{blue}{Our assumption that how individuals themselves were produced (i.e., by selfing or outcrossing) affects the level of inbreeding depression they suffer also represents a major simplification}. In reality, the history of consecutive generations of inbreeding in each individual's lineage will influence the severity of inbreeding depression they experience, particularly when inbreeding depression is caused primarily by recessive deleterious mutations. It would be an interesting and feasible extension of our modeling framework to expand the individual state space to include selfing cohorts (i.e., first generation selfing, second generation selfing, etc.), as in the models of \citet{kelly1999response,kelly2007mutation}, thereby enabling a more biologically plausible approach to modelling self-fertilization. \textcolor{blue}{Likewise, we implemented a simple "fixed prior selfing" model, where all individuals in the population self-fertilize at the same rate, which is constant over time. In reality, environmental stochasticity will likely influence the population selfing rate and therefore both the expression of inbreeding depression and the scope for SA polymorphism \citep{JordanConnallon2014,Olito-etal-2018}. In this scenario, a naive expectation might be that the geometric mean selfing rate over time would represent a more biologically meaningful parameter influencing the genetic and demographic outcomes of SA selection.}

@@

\cite{grieshop2017male} found that genotypes with sexually antagonistic alleles that are male-beneficial experience higher levels of inbreeding depression than genotypes with female beneficial SA alleles. This effect can easily be included in our model by making the inbreeding depression parameters, $\delta$ and $\delta_i$, a function of the genotype of the individual. Such an effect would further reduce the demographically viable polymorphic parameter space, and increase the bias in viable SA polymorphisms towards alleles with weaker selection in females.

>>>>

%To persist in the long term, a population must be able to adapt to its environment. Yet, the process of adaptation can be impeded by a variety of genetic and environmental factors, such as deleterious mutations \citep{Haldane1937}, \textcolor{blue}{rapid, unpredictable, or extreme changes} in environmental conditions \citep{Maynard-Smith1976, LandeShannon1996, OrrUnckless2008,Lyberger-etal-2021}, maladaptive gene-flow \citep{KirkpatrickBarton1997, BolnickNosil2007}, and genetic constraints \citep{ConnallonHall2018, MatthewsConnallon2019}. Genetic constraints on adaptation can arise as a consequence of conflicting selection and gene-flow between different classes of individuals, or more generally, when \textcolor{blue}{different fitness components are negatively correlated} \citep{CharlesworthHughes2000, ConnallonHall2018}. 
\textcolor{blue}{A population's ability to adapt to its environment and persist in the long term depends upon the genetic variation that it carries \citep{Fisher1930, Lewontin1975}. Yet, populations are heterogenous collections of individuals existing in a complex world where adaptation can be impeded by conflicting selection and gene-flow between different classes of individuals, or more generally, when different fitness components are negatively correlated \citep{Prout2000, CharlesworthHughes2000, ConnallonHall2018}.}

@@

Sexually antagonistic selection (abbreviated SA hereafter) \textcolor{blue}{is an important form of conflicting selection that occurs when beneficial alleles for one sex are deleterious when expressed in the other. Likewise, alleles with opposing fitness effects through male and female sex functions can cause SA selection in hermaphrodite populations, where individuals' fitness is determined jointly by their maternal and paternal reproductive success \citep{LloydWebb1986, WebbLloyd1986, Abbott2011, JordanConnallon2014}. SA selection appears to be} a common feature of sexually reproducing populations, and is thought to contribute to the maintenance of genetic variation and the evolution of sexual dimorphism and sex-related traits \citep{Kidwell1977, Lande1980, Rice1992, Charlesworth1999, RiceChippindale2001, ConnallonClark2012, BondurianskyChenoweth2009, Olito2019}. \textcolor{blue}{For instance, phenotypic selection studies commonly find sex-differences in directional selection \citep[e.g.,][]{DeLisle-etal-2018,Lewis-etal-2011,CoxCalsbeek2009,SinghPunzalan2018}, implying that many genetic variants underlying quantitative traits have SA fitness effects. Moreover, quantitative genetic studies frequently find negative cross-sex genetic correlations for fitness (\citealt{Chippendale-etal-2001,Delph-etal-2011}; reviewed in \citealt{ConnallonMatthews2019}), and polygenic signals of SA selection are evident in recent population genetic analyses of large genomic datasets \citep{RuzickaESEB2020, RuzickaConnallon2022}. Overall, both theory and current empirical data suggest that there is ample scope for the maintenance of SA \textcolor{blue}{genetic variation} in both dioecious \textcolor{blue}{(separate-sexed)} and hermaphrodite populations \citep{BondurianskyChenoweth2009, Abbott2011, RuzickaESEB2020, WangBarrett2020}.}

@@

\textcolor{blue}{While significant efforts have been made to predict when SA polymorphisms should be maintained and to detect SA genetic variation from empirical data, less attention has been paid to the demographic consequences of SA selection \citep{MatthewsConnallon2019}}. Most population genetic theory, including SA selection theory, models the relative fitness of genotypes in populations of constant (and often infinite) size. In reality, these genotypic fitnesses emerge from myriad processes acting throughout the life-history of individuals \textcolor{blue}{which link genotype distributions to population dynamics (e.g., \citealt {johnston2013life, merot2020balancing}). For over 50 years, the field of eco-evolutionary dynamics has sought to jointly model the genetic and demographic processes involved in evolutionary change. Theoretical models combining population genetics with density-dependent population growth \citep{roughgarden1971density} and stage-structured populations (consolidated in \citealt{Charlesworth_1994}) paved the way toward understanding the demographic consequences of allele frequency changes due to selection. Various extensions of these theories have been made to study life-history evolution \citep{orive1995senescence, orive2001somatic}, spatial structure \citep{ronce2000kin}, kin selection \citep{rousset2004genetic}, and the evolution of quantitative traits \citep[e.g.,][]{coulson2006putting, barfield2011evolution, childs2016evolution, orive2017effects}. This growing body of eco-evolutionary theory makes plain that demographic processes must be taken into account explicitly to fully understand how genetic variation for fitness is shaped and maintained in natural populations.}

@@

\textcolor{blue}{One particularly important limitation of our main analysis is that we modeled density-independent growth. Of course, in natural populations density-dependent processes like resource competition or competition for breeding sites will often prevent populations from growing (or shrinking) exponentially. In the simple density-dependent version of our model presented in \ref{App:DensDep}, we included density-dependence acting on the survival of all life history stages equally, and reanalysed our case study of \textit{M. guttatus}. The density-dependent version of the model predicts equilibrium population densities rather than population intrinsic growth rates (see fig.~ \ref{fig:mimulusFig_dd}). This brief example shows that the shape of the polymorphic region is unchanged by this form of density-dependence, and that the shape of the population density surface in fig.~\ref{fig:mimulusFig_dd} is very similar to that for population growth rate in fig.~\ref{fig:mimulusFig}. The similarity between the density-independent and -dependent model predictions underscores an important point: density-dependence that acts (st)age-independently does not alter selection gradients \citep{mylius1995evolutionarily, caswell2017}.}

\textcolor{blue}{In many natural populations, however, density-dependence will not have the same effect on the survival of individuals of all ages or developmental stages. For example, competition for breeding sites will cause a reduction in fecundity rather than survival. In plants, density-dependent competition for soil nutrients will likely influence seed germination rates differently than seedling growth rates or flower/fruit production in adults \citep{AntonovicsLevin1980}. These types of age- or stage-specific density-dependence will change the selection gradients, and therefore the shape of polymorphic parameter space as well as the equilibrium population densities predicted by the model. \citet{de2020matrix} discuss some of the possible dynamical outcomes of a density-dependent genetic model for \textit{Tribolium} seed beetles but a thorough analysis of more complicated forms of density-dependence for the maintenance of SA polymorphisms remains an important avenue for future work.}

@@

\textcolor{blue}{Our assumption that how individuals themselves were produced (i.e., by selfing or outcrossing) affects the level of inbreeding depression they suffer also represents a major simplification}. In reality, the history of consecutive generations of inbreeding in each individual's lineage will influence the severity of inbreeding depression they experience, particularly when inbreeding depression is caused primarily by recessive deleterious mutations. It would be an interesting and feasible extension of our modeling framework to expand the individual state space to include selfing cohorts (i.e., first generation selfing, second generation selfing, etc.), as in the models of \citet{kelly1999response,kelly2007mutation}, thereby enabling a more biologically plausible approach to modelling self-fertilization. \textcolor{blue}{Likewise, the simple 'fixed prior selfing' model we implemented assumes constant selfing rates among individuals over time. In reality, environmental stochasticity will influence the population selfing rate and therefore both the expression of inbreeding depression and the scope for SA polymorphism \citep{JordanConnallon2014,Olito-etal-2018}. A naive expectation might be that some form of weighted geometric mean selfing rate would represent a more biologically meaningful parameter influencing the genetic and demographic outcomes of SA selection in more complicated 'selfing environments' \citep[e.g., ][]{ConnallonSharmaOlito2019}.}

@@

Finally, \cite{grieshop2017male} found that genotypes with SA alleles that are male-beneficial experience higher levels of inbreeding depression than genotypes with female beneficial SA alleles. This effect can easily be included in our model by making the inbreeding depression parameters, $\delta$ and $\delta_i$, a function of the genotype of the individual. Such an effect would further reduce the demographically viable polymorphic parameter space, and increase the bias in viable SA polymorphisms towards alleles with weaker selection in females.
